### 2010-04-01 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: proof-read section 4, levitation

In this section, we will fulfil our promises. First, we will build
finite sets. Second, we will implement the code of the $\Desc$
universe.  Despite being a perilous pedagogical exercise for the
authors, we hope to convey to the reader the dizzy feeling of
levitation, without the falling.

@@

Let us consider the consequences. First, we discover that the type
theory does not need to be extended with the type former $\EnumU$, the
constructors $\NilE$ and $\ConsE$. We have just defined $\EnumU$
above, by fixpoint over the pattern functor denoted by
$\ListD$. $\NilE$ and $\ConsE$ are simply the $\ListNil$ and
$\ListCons$ constructors of lists.

Another interesting consequence concerns the $\spi{\!}{\!}$ operator:
it can actually be implemented with the generic $\F{induction}$
principle. From an implementation perspective, this means that this
operator is simply implemented in the dependently-typed target
language, as any other function.

@@

We shall now fulfil our promises for $\Desc$. We are going to
implement the codes of the universe of descriptions. In and by itself,
the system of codes, $\Desc$, is nothing but a datatype. We are in
the same situation as with $\EnumU$: we ought to be able to describe
the codes of $\Desc$ in $\Desc$ itself. Hence, this code would be a
first-class citizen, born with the standard, generic equipment of
datatypes.

@@

For $\DUnit$, we immediately reach the end of the description. For
$\DIndx{\!}$, there is a single recursive argument. The case of
$\DSigma{\!}{\!}$ is problematic. Recall the specification of
$\DSigma{\!}{\!}$:

@@

Note that $\DIndx{\!}$ is isomorphic to $\DHindx{\Unit}{\!}$. However,
we tolerate this duplication. Indeed, unlike its counterpart,
$\DIndx{\!}$ is first-order. Both codes are \emph{extensionally} the
same but, in practice, they will be dealt with rather
differently. While definitional equality can cope with first order
objects, the functional presentation introduced by $\DHindx{\!}{}$ is
unlikely to be amenable to a purely definitional treatment.

As shown in Figure~\ref{fig:desc-levitate}, we are now able to
describe our universe of datatypes.  The $\DUnit$ and $\DIndx{\!}$
cases remain unchanged, as expected. We successfully describe the
$\DSigma{\!}{}$ case, by a simple appeal to the higher-order induction
on $\V{S}$. The $\DHindx{\!}{\!}$ case consists in packing the $\V{H}$
in $\Set$ with a recursive argument.

At a first glance, we have achieved our goal. We have described the
codes of the universe of description. Taking the fixpoint of this
object gives us $\Desc$, up to isomorphism. We have implemented our
code of datatypes as an object levitating inside itself. However,
this levitation operation, just as any magic trick, relies on an
invisible cable. Let us reveal it, hence finishing our implementation.

@@

Our reader might be slightly confused to learn that the trick is
visible in the definition of $\DescD$
(Fig.~\ref{fig:desc-levitate}). Or rather, it is made invisible by
careful usage of type propagation. Indeed, let us try to elaborate
this term down to the low-level type theory. The $\EnumT{[ \ldots ]}$
construct elaborates to a finite set $e$ in $\EnumU$, inhabited by the
codes. Then, we can type-check the case definition, between square
brackets $[ \ldots ]$. This term $\V{f}$ is pushed into the type
$\EnumT{\V{e}} \To \Mu{\DescD}$. This corresponds to a finite function
definition, as presented in Figure~\ref{fig:type-checking}.

@@

\[
\PLAM{\V{x}}{(\EnumT{\V{e}})} \switch{\V{e}}{(\LAM{\_} \Mu{\DescD})}{\V{\pi^f}}{\V{x}}
\]

But this is quite problematic. We are still in the process of
constructing $\DescD$, and the motive of $\F{switch}$ is abruptly
begging for this very same $\DescD$. Despite our willingness, we
cannot materialise such motive.

However, we perfectly know what the motive is. Consequently, we extend
the type theory with a special-purpose operator:

@@

The entire work of the magician stands here, in this extension. Note
that the developer do not have to decide between the two variants of
$\F{switch}$. This is dealt with by the type propagation system
(Fig.~\ref{fig:type-checking}), extended by the following inference
rule:

@@

This concludes our levitation work. Beyond its pedagogical value, this
exercise has several practical outcomes. First of all, it reveals that
the $\Desc$ universe is just plain data. Just as any piece of data, it
can therefore be inspected and manipulated. Moreover, it is expressed
in the $\Desc$ universe. As a consequence, it is equipped, for free,
with an induction principle. So, our ability to inspect and program
with $\Desc$ is not restricted to the meta-language: we now have all
the necessary equipment in the target language to \emph{program} over
datatypes. In this setting, \emph{generic programming is just 
  programming}.

@@

The catamorphism is an induction on the description $\V{D}$, with a
non-dependent motive targeting $\V{T}$. Given a node $\V{xs}$ and the
induction hypotheses, the method ought to build an element of
$\V{T}$. Provided that we know how to make an element of
$\descop{\V{D}}{\V{T}}$, this step will be performed by the algebra
$\V{f}$. Let us take a look at this jigsaw:

@@

Filling the sub-goal with $\apply\: \V{D}\: \V{X}\: \V{Y}\:
\V{\sigma}$ completes the implementation. To sum up, we have
implemented the free monad construction for an arbitrary tagged
description. This gives the developer the ability, for any datatype,
to extend it with a notion of variable. Then, we have equipped this
structure with the corresponding monadic operation, \bind\ and
\return. We have shown how to implement \bind, for any free
monad. This construction is an example of type-indexed
datatype~\cite{hinze:generic-haskell}, as found in Generic Haskell:
from a datatype, we built a new datatype and equip it with its
structure.

>>>>

In this section, we will fulfil our promises. We first implement
finite sets. Then, we implement the codes of the $\Desc$ universe.
Despite being a perilous pedagogical exercise for the authors, we hope
to convey to the reader the dizzy feeling of levitation, without the
falling.

@@

Let us consider the consequences. First, we discover that the type
theory does not need to be extended with the type former $\EnumU$, the
constructors $\NilE$ and $\ConsE$. $\EnumU$ is defined above, as the
fixpoint of $\ListD~\UId$. $\NilE$ and $\ConsE$ are simply the type
constructors of lists, $\ListNil$ and $\ListCons$.

Another interesting consequence concerns the $\spi{\!}{\!}$ operator:
it can actually be implemented with the generic $\F{induction}$
principle. From an implementation perspective, this means that this
operator is implemented in the dependently-typed target language, as
any other function.

@@

We shall now fulfil our promise concerning $\Desc$. We are going to
implement the codes of the universe of descriptions. In and by itself,
the codes, $\Desc$, is nothing but a datatype. We are in the same
situation as with $\EnumU$: we ought to be able to describe the codes
of $\Desc$ in $\Desc$ itself. Hence, this code would be a first-class
citizen, born with the standard, generic equipment of datatypes.

@@

For $\DUnit$, we immediately reach the end of the description. For
$\DIndx{\!}$, there is a single recursive argument. Describing
$\DSigma{\!}{\!}$ is problematic. Recall the specification of
$\DSigma{\!}{\!}$:

@@

Note that $\DIndx{\!}$ is isomorphic to $\DHindx{\Unit}{\!}$. However,
we tolerate this duplication. Indeed, unlike its counterpart,
$\DIndx{\!}$ is first-order. Both codes are \emph{extensionally} the
same but, in practice, they will be dealt with rather
differently. While definitional equality can cope with first-order
objects, the functional presentation introduced by $\DHindx{\!}{}$ is
unlikely to be amenable to a purely definitional treatment.

As shown in Figure~\ref{fig:desc-levitate}, we are now able to
describe our universe of datatypes.  The $\DUnit$ and $\DIndx{\!}$
cases remain unchanged, as expected. We successfully describe the
$\DSigma{\!}{}$ case, by a simple appeal to the higher-order induction
on $\V{S}$. The $\DHindx{\!}{\!}$ case consists in packing an $\V{H}$
in $\Set$ with a recursive argument.

At a first glance, we have achieved our goal. We have described the
codes of the universe of description. Taking the fixpoint of this
object gives us $\Desc$, up to isomorphism. We have implemented the
codes of datatypes as an object levitating inside itself. However,
this levitation operation, just as any magic trick, relies on an
invisible cable. Let us reveal it, hence finishing our implementation.

@@

Our reader might be slightly confused to learn that the trick is
visible in the definition of $\DescD$
(Fig.~\ref{fig:desc-levitate}). Or rather, it is made invisible by
careful usage of type propagation. Indeed, let us try to elaborate
this term down to the low-level type theory. The $\EnumT{[ \ldots ]}$
construct elaborates to a finite set $\V{E}$ in $\EnumU$, inhabited by
the codes. Then, we can type-check the case definition, between square
brackets $[ \ldots ]$. This term $\V{f}$ is pushed into the type
$\EnumT{\V{E}} \To \Mu{\DescD}$. This corresponds to a finite function
definition, as presented in Figure~\ref{fig:type-checking}.

@@

\[
\PLAM{\V{x}}{(\EnumT{\V{E}})} \switch{\V{E}}{(\LAM{\_} \Mu{\DescD})}{\V{\pi^f}}{\V{x}}
\]

But this is quite problematic. We are still in the process of
constructing $\DescD$, and the motive of $\F{switch}$ is abruptly
begging for this very same $\DescD$. Despite our willingness, we
cannot materialise such motive. However, we perfectly know what the
motive is. Consequently, we extend the type theory with a
special-purpose operator:

@@

The entire work of the magician stands here, in this extension. Note
that the developer do not have to chose between the two variants of
$\F{switch}$. This is dealt with by the type propagation system
(Fig.~\ref{fig:type-checking}), extended with the following inference
rule:

@@

This concludes our levitation work. Beyond its pedagogical value, this
exercise has several practical outcomes. First of all, it reveals that
the $\Desc$ universe is just plain data. As any piece of data, it can
therefore be inspected and manipulated. Moreover, it is expressed in
the $\Desc$ universe. As a consequence, it is equipped, for free, with
an induction principle. So, our ability to inspect and program with
$\Desc$ is not restricted to the meta-language: we now have all the
necessary equipment in the target language to \emph{program} over
datatypes. In this setting, \emph{generic programming is just
  programming}.

@@

The catamorphism is defined by induction on the description $\V{D}$,
with a non-dependent motive targeting $\V{T}$. Given a node $\V{xs}$
and the induction hypotheses, the method ought to build an element of
$\V{T}$. Provided that we know how to make an element of
$\descop{\V{D}}{\V{T}}$, this step will be performed by the algebra
$\V{f}$. Let us take a look at this jigsaw:

@@

Filling the sub-goal with $\apply\: \V{D}\: \V{X}\: \V{Y}\:
\V{\sigma}$ completes the implementation. To sum up, we have
implemented the free monad construction for an arbitrary tagged
description. This gives the developer the ability, for any datatype,
to extend it with a notion of variable. Then, we have equipped this
structure with the corresponding monadic operation, \bind\ and
\return. This construction is an example of type-indexed
datatype~\cite{hinze:generic-haskell}, as found in Generic Haskell:
from a datatype, we built a new datatype and equip it with its
structure.
